---
title: "Practical Machine Learning- Course Project"
author: "Rakshith"
date: "April 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Practical Machine learning; Course Project

##1.Overview
This Course Project Report is part of the Practical Machine learning Coursera course submitted for peer review, The report is built using knitr and publised in html format.
The goal of this project is to predict the manner in which 6 participants performed some exercise as described below, This is the "classe" variable in the training set. The machine learning algorithm described here is applied to the 20 test cases available in the test data and the predictions are submitted in appropriate format to the Course Project Prediction Quiz for automated grading.

##2.Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3xsbS5bVX

##3.Data Extraction, Cleansing and Exploratory Data Analysis
###3.1 Data Overview
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from http://groupware.les.inf.puc-rio.br/har. Full source:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. "Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)". Stuttgart, Germany: ACM SIGCHI, 2013.

My special thanks to the above mentioned authors for being so generous in allowing their data to be used for this kind of assignment.

A short description of the datasets content from the authors' website:

"Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)."

###3.2 Environment Set Up

```{r echo=TRUE}
#clear some memory and load all required libraries
rm(list=ls())   
suppressWarnings(suppressMessages(library(knitr)))
suppressWarnings(suppressMessages(library(caret)))
suppressWarnings(suppressMessages(library(rpart)))
suppressWarnings(suppressMessages(library(rpart.plot)))
suppressWarnings(suppressMessages(library(rattle)))
suppressWarnings(suppressMessages(library(corrplot)))
suppressWarnings(suppressMessages(library(gbm)))

set.seed(1234)
```

###3.2 Data Preparation and Cleansing
Lets download the data from the Url provided and split the training data into Training(70%) and Testing set(30%) for validation. Testing set will be used for evaluating the result of the quiz.


```{r echo=TRUE}
trainURL="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL="http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training<-read.csv(url(trainURL))
testing<-read.csv(url(testURL))
#split training data into training and test data set
inTrain<-createDataPartition(training$classe,p=0.7,list=FALSE)

trainSet<-training[inTrain,]
testSet<-training[-inTrain,]
dim(trainSet)
dim(testSet)
```
Seems like there are 160 variables and most of them have NA values which could be eliminated by identifying NZV(near zero variance) variables.
```{r echo=TRUE}
 NZV<-nearZeroVar(trainSet)
trainSet=trainSet[,-NZV]
testSet=testSet[,-NZV]
dim(trainSet)

#Remove variable with 95% NA values
MostNA <- sapply(trainSet, function(x) mean(is.na(x))) > 0.95
trainSet = trainSet[,MostNA==FALSE]
testSet = testSet[,MostNA==FALSE]
dim(trainSet)

#Remove columns 1:5 as they are mostly ID columns and not a good predictor
trainSet=trainSet[,-(1:5)]
testSet=testSet[,-(1:5)]
```

##4 Prediction Model Building
We will be applying 3 different models on the training set 1.Decision Trees 2.Generalized Boosted Model 3.Linear Discriminate Analysis and pick the best one when applied to test set.

###4.1 Decision Trees 
```{r echo=TRUE}
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=trainSet, method="class")

# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=testSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, testSet$classe)
confMatDecTree

# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))

```

###4.2 Generalized Boosted Model

```{r echo=TRUE}
set.seed(12345)

controlGBM=trainControl(method="repeatedcv", number = 4,repeats = 1)
modFitGBM=train(classe ~ .,data=trainSet,method="gbm",trControl=controlGBM,verbose=FALSE)
modFitGBM$finalModel

#Prediction on test set
predictGBM=predict(modFitGBM, newdata=testSet)
confMatGBM <- confusionMatrix(predictGBM, testSet$classe)
confMatGBM
# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))

```

###4.3 Linear Discreminate Analysis
```{r echo=TRUE}
set.seed(12345)
modFitLDA=train(classe ~ .,data = trainSet, method="lda")

#Prediction on test set
predictLDA=predict(modFitLDA,newdata = testSet)
confMatLDA=confusionMatrix(predictLDA,testSet$classe)
confMatLDA

# plot matrix results
plot(confMatLDA$table, col = confMatLDA$byClass, 
     main = paste("LDA - Accuracy =", round(confMatLDA$overall['Accuracy'], 4)))

```

##5 Final Model Selection
Overall accuracy of the 3 models applied are as follows 

1.Decision Trees: 72.8% 
2.GBM: 98.74 
3.LDA: 71.84% 

GBM seems to be providing the best accuracy hence we will be applying this model to the testing set.

```{r echo=TRUE}
predictTest=predict(modFitGBM,newdata = testing)
predictTest
```








